Hilbert space methods to approximate Gaussian processes using Stan

### 2019-03-12 Paper/manuscript/manuscript_05.tex
some edits in intro

Gaussian processes (GPs) are flexible statistical models for specifying probability distributions over multi-dimensional non-linear functions \citep{rasmussen2006gaussian,neal1997monte}. Their name stems from the fact that any finite set of function values is jointly distributed as a multivariate normal. GPs themselves are defined by two functions, a mean and a covariance function, with the latter being arguably the more important one. The covariance function encodes our prior assumptions about the functional relationship, such as continuity, smoothness, periodicity and scale properties. GPs not only allow for non-linear effects but can also implicitly handle interactions between covariates. Different types of covariance functions can be used, and they may be combined for further increased flexibility. Due to their generality and flexibility, GPs are of broad interest across machine learning and statistics \citep{rasmussen2006gaussian,neal1997monte}. Among others, they find application in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, as well as Bayesian optimization and probabilistic numerics \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}.

Given $n$ observations in the data, the computational comlexity and memory requirements of exact GP implementation in general scale as $O(n^3)$ and $O(n^2)$, respectively. This would limit their application to rather small data sets of a few tens of thousands observations at most. The problem becomes especially severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed, which we will briefly review in the following.

Sparse GPs are based on low-rank approximations of the covariance matrix. The low-rank approximation with $m \ll n$ {\it inducing points} implies reduced memory requirements of $O(nm)$ and corresponding computational complexity of $O(nm^2)$.
A unifying view on sparse GPs based on approximate generative methods
is provided in \cite{quinonero2005unifying}, while a general review
can be found in \cite{rasmussen2006gaussian}.
% More recent developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.

An alternative class of low-rank approximations is based on forming a basis function approximation with $m \ll n$ basis functions. The basis functions are usually presented explicitly, but can also be used to form a low rank covariance matrix approximation. Common basis function approximations rest on the spectral analysis and series expansions of Gaussian processes \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}.
Sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving}. Recently, \cite{hensman2017variational} presented a variational Fourier feature approximation for Gaussian processes that was derived for the Mat{\'e}rn class of kernels. Another related method for approximating kernels relies on random fourier features \citep{rahimi2008random,rahimi2009weighted}.
Further, certain spline smoothing basis functions are equivalent to GPs with certain covariance functions \citep{wahba1990spline}, but not all spline models are GPs (e.g., sometimes splines are not marginalization consistent).

@@

In this paper we focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Basis function approaches behave computationally like linear models, which is an attractive property in modular probabilistic programming models where there is a big benefit id approximation specific computation is simple (Paul: What do you mean with the the last part of the sentence?). The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost when only changing the covariance function parameters is $O(mn+m)$ (or even $O(m^3)$ if marginalization can be used). This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks (cite!). This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation. The discussed basis function approach also provides an easy way to apply the non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which helps to further improve MCMC efficiency.

>>>>

Gaussian processes (GPs) are flexible statistical models for specifying probability distributions over multi-dimensional non-linear functions \citep{rasmussen2006gaussian,neal1997monte}. Their name stems from the fact that any finite set of function values is jointly distributed as a multivariate Gaussian. GPs are defined by a mean and a covariance function. The covariance function encodes our prior assumptions about the functional relationship, such as continuity, smoothness, periodicity and scale properties. GPs not only allow for non-linear effects but can also implicitly handle interactions between covariates. Different types of covariance functions can be combined for further increased flexibility. Due to their generality and flexibility, GPs are of broad interest across machine learning and statistics \citep{rasmussen2006gaussian,neal1997monte}. Among others, they find application in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, as well as Bayesian optimization and probabilistic numerics \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}.

Given $n$ observations in the data, the computational complexity and memory requirements of exact GP implementation in general scale as $O(n^3)$ and $O(n^2)$, respectively. This would limit their application to rather small data sets of a few tens of thousands observations at most. The problem becomes more severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed.

Sparse GPs are based on low-rank approximations of the covariance matrix. The low-rank approximation with $m \ll n$ {\it inducing points} implies reduced memory requirements of $O(nm)$ and corresponding computational complexity of $O(nm^2)$.
A unifying view on sparse GPs based on approximate generative methods
is provided in \cite{quinonero2005unifying}, while a general review
can be found in \cite{rasmussen2006gaussian}. \citet{Burt+Rasmussen+vanderWilk:2019} show that for regression with normally distributed covariates in $D$ dimensions and using the squared exponential covariance function, $M=O(\log^DN)$ is sufficient for accurate approximation.
% More recent developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.

An alternative class of low-rank approximations is based on forming a basis function approximation with $m \ll n$ basis functions. The basis functions are usually presented explicitly, but can also be used to form a low rank covariance matrix approximation. Common basis function approximations rest on the spectral analysis and series expansions of Gaussian processes \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}.
Sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving}. Recently, \cite{hensman2017variational} presented a variational Fourier feature approximation for Gaussian processes that was derived for the Mat{\'e}rn class of kernels. Another related method for approximating kernels relies on random fourier features \citep{rahimi2008random,rahimi2009weighted}.
Further, certain spline smoothing basis functions are equivalent to GPs with certain covariance functions \citep{wahba1990spline,Furrer+Nychka:2007}.
% , but not all spline models are GPs (e.g., some spline models are not marginalization consistent).

@@

In this paper we focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Basis function approaches behave computationally like linear models, which is an attractive property in modular probabilistic programming models where there is a big benefit if approximation specific computation is simple (Paul: What do you mean with the the last part of the sentence?). The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost when only changing the covariance function parameters is $O(mn+m)$.% (or even $O(m^3)$ if marginalization can be used).
This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks (cite!). This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation. The basis function approach also provides an easy way to apply the non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.
